---
title: "kde"
author: "Shawn Laffan"
date: "27 March 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries

Load libraries

Note that this process has been developed using this as a reference:
https://mgimond.github.io/Spatial/point-pattern-analysis-in-r.html


```{r}
library (rgdal)
library (sp)
# library (spdep)
library (sf)
library (maptools)
library (raster)
library (spatstat)


```

## Load your data

Remember to use forward slashes for directory separators.

If you are using a shapefile then set ```source_db``` to equal ```source_dir```.  If using a geodatabase then specify the geodatabase directory name.  If using a geopackage then use the .gpkg file name.

Directories are called "folders" on windows.

Note that rgdal can read geodatabase feature classes and tables, but cannot wite to them. We will use the geopackage format for outputs by default as they can be read by both QGIS and ArcGIS, whle also allowing for long field names.


```{r}

source_dir   = "C:/shawn/lab_data/Aus_quakes"
# source_db    = paste(source_dir, "census_subset.gdb", sep="/")
source_db    = source_dir
source_layer = "quakes_gt0"  #  no .shp extension needed
intens_field = "MAG"  #  intensity/pop count field, set to NULL to ignore

output_dir    = source_dir
output_driver = "GeoTIFF"
output_raster_pfx = paste("kde", intens_field, sep="_")
output_overwrite = FALSE


data = readOGR(dsn = source_db, layer = source_layer)

no_marks = FALSE
if (is.null (intens_field)) {
  #  should check if this exists, but we save to new data anyway.
  intens_field = "___DEFAULT_INTENS_FLD___"
  data[[intens_field]] = 1
  no_marks = TRUE
}


if (!is.projected(data)) {
  message ("Data are not in a projected coordinate system, reprojecting")
  #  project here
  #  epsg 3577 is Australian Albers, see epsg.io for more proj.4 strings
  #  e.g mollweide is (from http://epsg.io/54009)
  #  "+proj=moll +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs"
  proj_code = "+init=epsg:3577"
  data = spTransform (data, proj_code)
}

data_sf = st_as_sf (data)  #  coerce data into sf class to check geometry type
geom_type = class(data_sf$geometry)
is_poly_data  = any (grepl ("POLY", geom_type))
is_point_data = any (grepl ("POINT", geom_type))

if (!(is_poly_data || is_point_data)) {
  message ("Input data are neither points nor polygons, this might not work")
}
if (is_poly_data) {
  message ("Input data are polygon, will use centroids.  Be warned.")
}

data_ppp = as (data, "ppp")
if (no_marks) {
  marks(data_ppp) = NULL
}

```


Plot the data if there are not too many features.   Otherwise it takes a very long time. 

```{r}
#  plotting can take a long time
if (nrow(data) < 3000) {
  spplot (data, zcol=c(intens_field))
  # spplot (data, zcol=indep_field)
} else {
  message ("too many features to plot by default, modify the code to show more")
}

summary (data[c(intens_field)])

```


Now we run a KDE analysis.


```{r}

distances = c(100000, 20000000, 300000, 400000, 500000)
distances = c(100000, 200000)

#  set the analysis window
#  CLUNK
Window(data_ppp)

wt_func = "epanechnikov"  #  matches ArcGIS

kde.results = list();

i = 0
for (target_dist in distances) {
  i = i + 1
  message ("Running bandwidth ", target_dist)
  
  kde.res <- density(
    data_ppp, 
    sigma   = target_dist, 
    kernel  = wt_func, 
    weights = data[[intens_field]]
  )
  #  avoid scientific notation in names
  label = format(target_dist, scientific=FALSE)
  kde.results[[label]] = kde.res

}
```



Now we need to save the results to a set of geotiffs, one per analysis.
We will use the geotiff format for the spatial data.
This will fail if you have the output open in a GIS.

```{r}

distances = names(kde.results)

i = 0
for (model in kde.results) {
  i = i + 1
  distance = format (distances[i], scientific=FALSE)
  # is.character(distance)
  df = model$SDF

  output_name = paste (output_raster_pfx, distance, sep="_")
  output_name = paste0 (output_name, ".tif")
  output_full_path = paste (output_dir, output_name, sep="/")

  message (
    paste (
      "Writing",
      output_full_path
    )
  )

  writeRaster (
    raster(model),
    output_full_path, 
    overwrite = overwrite
  )
  
}


```

