---
title: "BEAST with spatial data"
output: html_notebook
---

Load needed libraries.  Install if necessary.

```{r}

pack_list = c("terra", "Rbeast", "stringi", "svDialogs", "ncdf4", "lubridate", "methods")
install_list = pack_list[!(pack_list %in% installed.packages()[,"Package"])]
if(length(install_list)) install.packages(install_list)

for (p in pack_list) {
  library(p, character.only = TRUE)
}

if (!hasMethods("toMemory", package="terra")) {
  #  update terra
  install.packages("terra")
  library("terra")
}


```

Set up some functions 

```{r}

beast_modis = function (raster, start_time = NULL, ...) {

  #  work on a copy in case it is not in memory
  #  then the copy should be GC'd when that is called
  tmp_ras = raster
  tmp_ras = toMemory(tmp_ras)   # To use beast, make sure all the data is read into memory
  dims    = dim(tmp_ras)

  #  generate time axis if needed
  #  assumes form 2024-01-24 somewhere in band name
  if (all (is.na(time(tmp_ras)))) {
    dates = names(tmp_ras)
    pattern = r"(\b\d{4}-\d{2}-\d{2}\b)"
    m = regexpr(pattern, dates)
    dates = regmatches(dates, m)
    time(tmp_ras) = strptime(dates, "%Y-%m-%d")
  }

  # Y = values(tmp_ras)
  #dim(Y)   = dims[c(2,1,3)]    # Assign column-major dim expected by Rbeast
  #  avoid the need to transpose
  Y = as.matrix(tmp_ras, wide=TRUE)
  dim(Y)   = dims[c(1,2,3)]  #  could just use dims directly...

  metadata = list(
    time             = time(tmp_ras),
    isRegularOrdered = FALSE,    # IRREGULAR input
    whichDimIsTime   = 3,        # 437 is the ts length, so set it to '3' here.
    # time$datestr     = datestr,  # date info is contained in the file names
    # time$strfmt      = 'LT05_018032_20080311.yyyy-mm-dd',
    #deltaTime        = 16/365,     # MODIS data are 16 days
    deltaTime        = 1/12,
#    period = 32/365
    period = 1
    #period           = 16/365
    #startTime        = ifelse (is.null(start_time, time(tmp_ras)[1], as.Date(start_time)))
  )

  #  minimal for now
  extra = list (
    numThreadsPerCPU = 3,
    numParThreads    = 30
  )

  #browser()
  tmp_ras = NULL
  gc()


  o = beast123 (Y, metadata = metadata, extra = extra, ...)
  
  #  pack in some extra info - underhanded but works for now
  o$ext = as.vector(ext(raster)) #  ext objects do not survive serialisation otherwise
  o$crs = crs(raster)

  gc()
  return (o)
}

#  convert a coordinate copied from ArcGIS to a format usable with terra
#  requires the extent be stored on the rbeast object
coord2idx_rbeast = function (b, coord) {
  coord = stri_trim(coord)
  xy = stri_split_regex(coord, "\\s")[[1]][1:2]
  xy = stri_replace_all_fixed (xy, replacement="", pattern=",")

  x_hemi = stri_extract(xy[1], regex = "[NESW]")
  x = stri_replace_all_fixed(xy[1], replacement = "", pattern = x_hemi)
  x = as.numeric(x)
  if (x_hemi == "W") {
    x = x * -1
  }

  y_hemi = stri_extract(xy[2], regex = "[NESW]")
  y = stri_replace_all_fixed(xy[2], replacement = "", pattern = y_hemi)
  y = as.numeric(y)
  if (y_hemi == "S") {
    y = y * -1
  }

  extent = ext(b$ext)

  if (x < extent$xmin || x >= extent$xmax) {
    stop ("X coord is outside data set extent")
  }
  if (y < extent$ymin || y >= extent$ymax) {
    stop ("Y coord is outside data set extent")
  }
  
  ncol = b$ncols
  nrow = b$nrows
  
  c = 1    + floor ((x - extent$xmin) / (extent$xmax - extent$xmin) * ncol)
  r = nrow - floor ((y - extent$ymin) / (extent$ymax - extent$ymin) * nrow)

  res = c(r, c, use.names=FALSE)
  return (res)
}


plot_beast_modis_coord = function (b, coord, t=FALSE) {
  rowcol = coord2idx_rbeast(b, coord)
  
  
  #  transpose for beast coords
  if (t) {
    rc = rowcol[c(2,1)]
  }
  else {
    rc = rowcol[c(1,2)]
  }
  message ("Plotting image row ", rc[1], ", col ", rc[2])
  message ("Mean number of trend change points is ",  b$trend$ncp[rc[1],rc[2]])
  message ("Mean number of season change points is ", b$season$ncp[rc[1],rc[2]])
  plot (b[rc])

}

load_data = function (file = NULL, drivers=NULL) {
  if (is.null(file)) {
    file = file.choose()
  }
  r = rast(file, drivers=drivers)
  if (any (is.na(time(r)))) {
    dates = strptime(names(r), "%Y-%m-%d")
    if (any(is.na(dates))) {
      stop ("Some of the field names do not satisfy the date format requirement (yyyy-mm-dd)")
    }
    time(r) = as.Date(dates)
  }
  return (r)
}


beast_time_to_date = function (x) {
  #  this fails because the current month and day are assumed for %Y on its own
  #b$time_as_date = as.Date(as.character(floor(b$time)), format="%Y") + (b$time - floor (b$time))*365
  as.Date(paste(floor(x), ceiling((x - floor (x))*365), sep=""), format="%Y%j")
}

rasterise_beast = function (b) {
  if (is.null(b$ext)) {
    stop ("An extent must be set on the beast object")
  }
  
  b$time_as_date = lubridate::date_decimal (b$time)
  
  results = list ()
  trend_ncp = beastbit2raster (b, "trend", "ncp")
  
  for (component in c("trend", "season")) {
    for (subcomponent in names(b[[component]])) {
      if (!is.null (b[[component]][[subcomponent]])) {
        message (sprintf("%s: %s", component, subcomponent))
        label = sprintf ("%s_%s", component, subcomponent)
        results[[label]] = beastbit2raster(b, component, subcomponent)
      }
    }
  }
  for (component in c("R2", "RMSE", "sig2", "marg_lik")) {
    message (component)
    label = component
    results[[label]] = beastbit2raster(b, component)
  }
  
  
  results
}

beastbit2raster = function (b, component = "trend", subcomponent = "ncp", template=NULL) {
  if (class(b) != "beast") {
    stop ("Need an Rbeast result")
  }

  temporal_subcomponents = c(
    "cpOccPr", "Y", "SD", "order", 
    "slp",     "slpSD",
    "slpSgnPosPr", "slpSgnZeroPr",
    "amp",     "ampSD"
  )
  
  valid_components = c("trend", "season", "R2", "RMSE", "sig2", "marg_lik")
  if (!(component %in% valid_components)) {
    stop (cat ("component args must be in list: ", paste(valid_components, sep=", ")))
  }
  if ((component %in% c("trend", "season")) && is.null(b[[component]][[subcomponent]])) {
    stop (sprintf ("Cannot find b$%s$%s, or it is null", component, subcomponent))
  }

  ext = b$ext
  if (is.null(ext) && (is.null (template) || !hasMethod("ext", class(template)))) {
    stop ("No extent found. Need an Rbeast object with an ext slot, or a template object with an ext method")
  }
  ext = ext(ext) #  ensure we have an extent object

  crs = b$crs
  if (is.null(crs) && (is.null (template) || !hasMethod("crs", class(template)))) {
    stop ("No coord sys found.  Need an Rbeast object with a crs slot, or a template object with a crs method")
  }
  
  rasterise = function (data) {
    m = matrix(
      data,
      b$nrows[1],
      b$ncols[1],
      byrow = FALSE
    )
    r = terra::rast(m, extent=ext, crs=crs)
  }

  r = NULL
  
  if (component %in% c("trend", "season")) {
    subdata = b[[component]][[subcomponent]]
    dims = dim(subdata)
    if (length (dims) > 4) {
      message (sprintf ("Too many dims, skipping: %s, %s", component, subcomponent))
    }
    else if (length(dims) < 3) {
      r = rasterise (subdata)
      names(r) = sprintf ("%s_%s", component, subcomponent)
    }
    else if (length(dims) == 3) {
      nbands = dims[3]
      r = do.call("c", lapply(1:nbands, function(i){
        return (rasterise (subdata[,,i]))
      }))
  
      #  These should be temporal data.
      if (subcomponent %in% temporal_subcomponents && nbands == length(b$time)) {
        message (sprintf ("Temporal data for %s, %s", component, subcomponent))
        t = b$time_as_date  #  use pre-calculated if exists
        if (is.null(t)) {
          t = lubridate::date_decimal (b$time)
        }
        time(r) = t
      }

      names(r) = paste0 (
        sprintf ("%s_%s", component, subcomponent),
        formatC(1:nbands, flag="0", width=nchar(nbands))
      )

    }
    else {
      ci_names = c("lower", "upper")
      r = list()
      for (j in 1:2) {
        
        nbands = dims[3]
        rr = do.call("c", lapply(1:nbands, function(i){
          return (rasterise (subdata[,,i,j]))
        }))
  
        names(rr) = paste0 (
          sprintf ("%s_%s_%s", component, subcomponent, ci_names[j]),
          formatC(1:nbands, flag="0", width=nchar(nbands))
        )
        if (length(r) > 0) {
          r = append (r, rr)
        }
        else {
          r = rr
        }
      }
      
    }
  }
  else {
    subdata = b[[component]]
    dims = dim(subdata)
    r = rasterise (subdata)
    names(r) = component
  }
  
  r
}

export_beast_rasters = function (b, dir, prefix="", overwrite=FALSE) {
  list = rasterise_beast(b)
  message ("Exporting now")
  for (name in names(list)) {
    pfx = file.path (dir, paste0(prefix, name))
    outfile = paste0 (pfx, ".tif")
    message (outfile)
    r = list[[name]]
    r[r == Inf] = NA
    writeRaster(list[[name]], outfile, overwrite=overwrite)
    #  also dump netCDF for temporal data
    if (!anyNA(time(r))) {
      outfile = paste0 (pfx, ".nc")
      message (outfile)
      writeCDF(list[[name]], outfile, overwrite=overwrite)
    }
  }
  invisible(list)
}


```

Now load the file we want and display some of its properties, and the first band.

Make sure you load a geotiff or netcdf file with _ONLY ONE INDEX_, 
e.g. EVI or NDVI, not both.  The code does not handle more than that and weird 
things will happen.  

```{r}

fname = file.choose()

raster = rast(fname)

dim (raster)

time(raster)[1:6]

t1 = time(raster)[1]

plot (raster[[1]], main = as.character(t1))

```

Now run beast on the loaded data set.  This might take a while.

```{r}

beast_result = beast_modis (raster)

plot (beast_result)



```

Some interaction.  Rerun this step each time you want to plot a different location.

The coordinates to enter are obtained by right clicking on the map in ArcGIS
and choosing "Copy Coordinates".

In ArcGIS:

  Make sure the map's display uses the MODIS Sinusoidal coordinate system.
  Also that the map's "Display units" are in "meters".
  Both of these can be set under the map's properties (right click on the map name 
  in the table of contents).  Display units is under the General section, 
  the Coordinate Systems should be self-documenting.

  The display units can also be changed by clicking 
  the drop down to the right of the coordinate at the bottom of the map pane.

```{r}


coord_string = dlg_input("Enter a coordinate from ArcGIS in the MODIS coord sys")$res

if (length (coord_string)) {
  plot_beast_modis_coord (beast_result, coord = coord_string)
}


```

Now export the beast results to a set of rasters.  

```{r}

output_dir = svDialogs::dlgDir()$res

output_prefix = svDialogs::dlg_input(message = "Filename prefix", default="")$res

save_flag      = svDialogs::dlg_list  (choices = c("No", "Yes"), title = "Overwrite existing files?")$res
overwrite_flag = ifelse (length(save_flag) > 0 && save_flag == "Yes", TRUE, FALSE)

b_rasters = export_beast_rasters(beast_result, output_dir, output_prefix, overwrite_flag)


```
